\section{Related Work}
\label{sec:related}

We discuss the most closely related work
on the metatheory of unrefined and refined
type systems.


% \subsection{Metatheory for Refined Systems}

\mypara{Hybrid \& Contract Systems}
%
\citet{flanagan06} formalizes on paper
a monomorphic lambda calculus with
refinement types that differs from our \sysrf
in two ways.
%
%Our soundness formalization follows
First, in \citet{flanagan06}'s type checking
is hybrid: the developed system
is undecidable and inserts runtime
casts when subtyping cannot
be statically decided.
%
Second, the original system lacks polymorphism.
%
\citet{SekiyamaIG17} extended hybrid types
with polymorphism, but unlike \sysrf,
their system does not support semantic subtyping.
%
For example, consider a divide by zero-error.
The refined types for \ha{div} and 0 could
be given by
$\mathsf{div} :: \funcftype{\tint}{\breft{\tint}{n}{n \neq 0} \rightarrow {\tint}}$
and $0 :: \breft{\tint}{n}{n=0}$.
This system will compile \ha{div 1 0}
by inserting a cast on 0:
$\langle\breft{\tint}{n}{n=0} \Rightarrow \breft{\tint}{n}{n\neq 0}\rangle$,
causing a definite runtime failure
that could have easily been prevented
statically.
%
Having removed semantic subtyping,
the metatheory of~\citet{SekiyamaIG17}
is highly simplified.
%
% \mypara{Decidable Systems}
%
Static refinement type systems
(as summarized by~\citet{sprite})
usually restrict the definition
of predicates to quantifier-free
first-order formulae that can be
\emph{decided} by SMT solvers.
%
This restriction is not preserved
by evaluation that can substitute
variables with any value, thus
allowing expressions that cannot
be encoded in decidable logics, like
lambdas, to seep into the predicates of types.
%
In contrast, we allow predicates to be
any language term (including lambdas)
to prove soundness via preservation
and progress: our meta-theoretical
results trivially apply to systems that,
for efficiency of implementation, restrict
their source languages.
%
Finally, none of the above systems (hybrid,
contracts or static refinement types) come
with a machine checked soundness proof.

% \newtext{
  \mypara{Semantic Subtyping}
  %
  Semantic subtyping is not a unique feature of refinement types. 
  For example, \citet{Castagna02} use the set theoretic models of types 
  to decide subtyping. 
  \citet{Castagna05} present an algorithm that decides semantic subtyping 
  for a core calculus with functional types. 
  Like \sysrf, \citet{Castagna05} introduce a denotational interpretation of types 
  to break the circularity between the typing and subtyping relations.
  Unlike \sysrf, their system does not have polymorphism and, crucially, has no notion of dependency
  (no refinement type-style binder of arguments).
  %
  Moreover, their subtyping algorithm is different than our refinement based algorithm:
  it is neither type directed nor efficient (\ie it requires backtracking), and
  cannot be automated by an external SMT solver. 
% }

\mypara{Mechanizations of Refinement Types}
%
%Our soundness formalization follows
%the axiomatized implication relation
%of \citet{LehmannTanter} that decides
%subtyping (our rule \sBase) without
%formally connecting implication and
%expression evaluation.
%
\citet{LehmannTanter}'s \coq formalization
of a monomorphic, refined calculus
differs from \sysrf
in two ways.
%
First, their axiomatized implication, 
which is similar to our implication interface,
allows them to 
restrict the language of refinements
to decidable logics but provides no formal connection 
between subtyping and evaluation.
%
Instead, we also provide the denotational 
implementation of the implication interface, 
thus establish denotation soundness.
%We allow refinements to be arbitrary
%program terms and intend, in the future,
%to connect our axioms to SMT solvers
%or other oracles.
%
Second, \sysrf includes polymorphism,
existentials, and selfification
which are critical for % path- and
context-sensitive refinement typing,
but make the metatheory more challenging.
%
% \mypara{System FR}
%
\citet{kuncak-stainless} present System FR,
a polymorphic, refined language
with a mechanized metatheory
of about 30K lines of \coq.
%
Compared to our system, their notion
of subtyping is not semantic, but relies on
a reducibility relation.
%
For example, even though System FR
will deduce that \tpos is a subtype of
\tint, it will fail to derive that
$\tint \rightarrow \tpos$ is subtype
of  $\tpos \rightarrow \tint$
as reduction-based subtyping
cannot reason about contra-variance.
%
Because of this more restrictive notion
of subtyping, their mechanization
requires neither the indirection of
denotational soundness nor 
an implication proving oracle.
%
Further, System FR's support
for polymorphism is limited
in that it disallows refinements
on type variables, thereby
precluding many practically
useful specifications. %\RJ{check!}
%
% \newtext{
  Recently, \citet{10.1145/3546196.3550162}
  formalized a 
  refinement type system 
  as an embedding of refinement types in Agda.
  This system is verified 
  in a few thousand lines of Agda.
  %
  This formalism differs significantly from ours in that
  as an embedding it is built on top of a rich theorem prover 
  and cannot be used to refine some existing programming language.
  %
  Further, 
  it does not support higher-order
  functions, polymorphism, 
  semantic subtyping, neither be automated by an external solver 
  since soundness reduces to Agda's soundness.
% }
%
Finally, \citet{explicit} mechanize refinement types
with explicit proof terms in 15K lines of \lean code. 
They use a categorical, denotational semantics 
soundness statement, but their calculus 
by design supports neither semantic subtyping
nor polymorphism. 

\mypara{Metatheory in \lh}
\texttt{LWeb}~\cite{lweb} also used \lh 
to prove metatheory, %concretely 
the non-interference of $\lambda_{\text{LWeb}}$,
a core calculus that extends the LIO formalism with database access. 
The \texttt{LWeb} proof did not use refined data propositions, 
which were not present at development time,
%the time of its development, 
and thus it has two major weaknesses compared to our 
present development. 
First, \texttt{LWeb}
\textit{assumes} termination of $\lambda_{\text{LWeb}}$'s evaluation function; 
without refined data propositions metatheory can be developed only over 
terminating functions. 
This was not a critical limitation since non-interference was only 
proved for terminating programs. 
However, in our proof the requirement that evaluation of \sysrf terminates 
would be too strict. 
In our encoding with refined data propositions
such an assumption was not required. 
Second, the \texttt{LWeb} development is not constructive: 
the structure of an assumed evaluation tree is logically inspected 
instead of the more natural case splitting permitted only with 
refined data propositions. 
This constructive way to develop metatheories is more compact 
(\eg there is no need to logically inspect 
%the structure of the 
derivation trees) and akin to the standard 
meta-theoretic developments of constructive tools like 
\coq and \isabelle. 
